A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2018-02-08 paper.tex
adding suggested edits from Keywan

\texttt{aneris}, first made available in 2017, is available online at
\url{https://github.com/gidden/aneris} as a free and open-source Python software
library (approximately 2000 lines of code). The \texttt{aneris} software was
developed by the lead author whose contact information is shown on the title
page of this manuscript. Documentation for the \texttt{aneris} Python package,
including software requirements, is available
\href{http://mattgidden.com/aneris/}{online}.

@@

\glspl{iam} are tools used to understand the complex interactions between
energy, economy, land use, water, and climate systems. \glspl{iam} provide
global projections of systemic change by dividing the world into a number of
representative regions (typically 10 to 30), the definition of which is distinct
for each model \cite{krey_global_2014}. Results from \glspl{iam} are integral in
a number of international studies, which notably include projections of climate
and energy futures. Recently, the \gls{iam} community has developed scenarios
based on the \glspl{ssp} \cite{fujimori_ssp3:_2017, van_vuuren_energy_2017,
  calvin_ssp4:_2017, fricko_marker_2017, kriegler_fossil-fueled_2017} which
quantify a variety of potential global futures. The \glspl{ssp} are designed to
be used in research that include \gls{esm} simulations, climate impact,
adaptation and climate mitigation studies \cite{vuuren_new_2013}.

While \glspl{iam} are implemented in myriad ways\footnote{IAM models are
  numerous and have a long history in the scientific literature. Various IAMs
  have collaborated to produce
  \href{http://themasites.pbl.nl/models/advance/index.php/ADVANCE_wiki}{community
    IAM documentation} which readers can access for a full treatment of model
  implementation and features.}, including simulation and optimization, the core
inputs and outputs are similar across different models. Modeling teams
incorporate data on energy systems, land use, economics, demographics and
emissions sources and concentrations, among other data, in order to provide a
consistent starting point for future projections. The models then provide
estimates of future trajectories of these variables under various socio-economic
and technological assumptions as well as proposed policy constraints, e.g.,
targets for future \gls{ghg} emissions.

@@

When participating in intercomparison exercises in which a consistent historical
starting point is required (e.g., in \gls{cmip6}), model teams incorporate a single,
common historical data set through \textit{harmonization}. Harmonization
refers to the process of adjusting model results to match a selected
historical time series such that the resulting future trajectories are also
consistent with the original modeled results. In the emissions context, this
means that each individual combination of model region, model sector, and
emissions species must be harmonized. Depending on the total number of model
regions, sectors, and emissions species, this can require the selection of
thousands to tens-of-thousands of harmonization methods.

@@

The remainder of this paper describes the methodology and implementation of the
harmonization software \code{aneris} \cite{matthew_gidden_2017_802832}, written
in the Python programming language (detailed documentation is available
\href{http://mattgidden.com/aneris/}{online}). Section \ref{sec:meths} provides
a detailed description of the underlying mathematical components of
\code{aneris} as well as the procedural workflow. A case study of applying the
automated harmonization mechanism on two example \gls{iam} scenarios, one with
emissions growth and another with emissions mitigation, is presented in Section
\ref{sec:results}. Finally, the general effectiveness and potential future
improvements on the automated methodology is discussed in Section
\ref{sec:future}.

@@

The goals of any harmonization exercise are threefold: aligning model results in
the harmonization year to a common historical data source, faithfully
representing the original \gls{iam}’s internal consistency between the driver of
emissions (e.g. energy use) and emissions, and maintaining critical parameters
from the original scenario design. Any harmonization method achieves the first
goal by design. If the difference between the model base year and historical
values are small, considering the second and third goals leads to a method
choice that matches modeled drivers (e.g., a \code{ratio} method discussed in
Section \ref{sec:harm-methods}) and converges prior to the final model year. It
preserves the relationship between \gls{iam} output and emissions inventory in
the base year while also matching the original model output at some point in the
modeled time period. It furthermore maintains the consistency of the model's
usage of energy technology, volume of agricultural activities, and abatement
options with harmonized emissions trajectory.

@@

Accordingly, we have developed a \textit{decision tree} approach to
harmonization method choice, discussed in Section \ref{sec:tree}, in order to
balance each of these concerns and use cases in a robust and automated manner.

@@

\gls{iam} emission results are provided along temporal (normally half decade or
decade), spatial (i.e., model regions), gas species, and sectoral
dimensions. Each individual temporal trajectory, i.e., unique combinations of
regions ($r$), species ($g$), and sectors ($s$), must be harmonized to the
initial modeling period. Given a model trajectory, $m_{r, g, s}(t)$, historical
values, $h_{r, g, s}(t)$, and model base year, $t_i$, a harmonized trajectory
needs to be calculated. The \textit{harmonization quality} of a trajectory,
i.e., how well a given harmonization algorithm performs, depends on a number of
factors. Of chief import is the faithful representation of the original,
unharmonized, trajectory as well as the representation of negative trajectories
(i.e., if a trajectory becomes negative, both the timing and total magnitude
should be as close as possible) which are of critical importance for cumulative
\cotwo calculations.

@@

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent \gls{ceds} and \gls{luc}
historical data has been performed. Figure \ref{fig:cov} shows the distribution
of \gls{luc} $c_v$s and non-\gls{luc} $c_v$s as determined for historical data
aggregated to the model regions of 5 different \glspl{iam} involved in the
\gls{ssp} process: AIM-CGE \cite{fujimori_ssp3:_2017},
IMAGE\cite{van_vuuren_energy_2017}, GCAM4\cite{calvin_ssp4:_2017},
MESSAGE-GLOBIOM\cite{fricko_marker_2017}, and
REMIND-MAGPIE\cite{kriegler_fossil-fueled_2017}, each of which have varying
definitions of native model regions comprising different collections of
countries. Therefore, each data point comprising Figure \ref{fig:cov} represents
a realization of $c_v$ for a single combination of native model region, sector,
and emissions species\footnote{A full listing of all sectors and species is
  presented in the case-study discussion in Section \ref{sec:results}, Table
  \ref{tab:sp}}. A threshold value of $\tau_{c_v} = 20$ has been chosen based on
these observations as it optimally divides the two distributions. Importantly,
tails of the \gls{luc} and non-\gls{luc} overlap, thus there are both false
positives (~7\% of non-\gls{luc} trajectories) and false negatives (~10\% of
\gls{luc} trajectories). However, as any regional definition is model dependent
and thus any regional aggregation is possible an automated detection mechanism
is necessary.

@@

\begin{listing}[H]
\begin{minted}{python}
from aneris.tutorial import load_data
model, hist, driver = load_data()
for scenario in driver.scenarios():
    driver.harmonize(scenario)
harmonized, metadata = driver.harmonized_results()
\end{minted}
\caption{High-level user interaction with the \texttt{HarmonizationDriver} taken from the online \href{http://mattgidden.com/aneris/config.html}{tutorial}}
\label{lst:tutorial}
\end{listing}

@@

optional arguments:
  -h, --help            show this help message and exit
  --history HISTORY     Historical emissions in the base year.
  --regions REGIONS     Mapping of country iso-codes to native regions.
  --rc RC               Runcontrol YAML file (see
                        http://mattgidden.com/aneris/config.html for
                        examples).
  --output_path OUTPUT_PATH
                        Path to use for output file names.
  --output_prefix OUTPUT_PREFIX
                        Prefix to use for output file names.
\end{minted}
\caption{The \gls{cli} help provided by the \code{aneris} package.}
\label{lst:cli}
\end{listing}

@@

Users are able to control the harmonization process via a number of options
(with examples provided
\href{http://mattgidden.com/aneris/config.html}{online}). The primary mechanism
by which users control the process is by providing \textit{override} methods for
any combination of region and variable (i.e., sector and gas species).  In
practice, it may be possible that not all default methods chosen will provide
robust harmonized trajectories, especially if there is a significant difference
between historical and model values in the harmonization year, if there is
significant upward or downward movement in the model trajectory, or if there are
known discrepancies in sectoral definition between the \gls{iam} and historical data
source. In such cases, users can override default methods with their method of
choice and both the default method and override is reported in the resulting
metadata.

@@

In order to show a representative cross section of the performance of the
\code{aneris} harmonization procedure, we focus on the harmonization of results
of the \gls{iam} MESSAGE-GLOBIOM \cite{fricko_marker_2017}. Harmonization
results for two scenarios from the \gls{ssp} scenario library\footnote{We refer
  the reader to the broad literature discussing the SSPs, e.g.,
  \cite{vuuren_new_2013,oneill_scenario_2016, oneill_new_2013} for a more in
  depth discussion of the scenario architecture and design.} are presented
here. We use the SSP2-reference\cite{Riahi2017153,Rao2017346}, or ``middle of
the road'', scenario (referred to as SSP2-Ref) as an example because
MESSAGE-GLOBIOM is the marker scenario\footnote{The ``marker'' scenario concept
  is used to designate the archetype scenario used as a reference within each
  SSP scenario family. See \cite{vuuren_new_2013} and \cite{Riahi2017153} for a
  more lengthy description.} for this \gls{ssp}. This SSP2 scenario lies between
two \glspl{rcp}, 6 and 8.5, with a radiative forcing\footnote{Radiative forcing
  in this context is the energy imbalance at the top of the atmosphere caused by
  anthropogenic influences relative to a pre-industrial reference point. Higher
  radiative forcing leads to larger global changes, such as surface
  temperature. We refer the reader to \cite{IPCCWG1PhysicalStocker2013} for a
  more detailed discussion.} level of approximately 6.5 Wm$^{-2}$. We
additionally present the results for the SSP2-based mitigation scenario leading
to a radiative forcing of 4.5 Wm$^{-2}$ (referred to as SSP2-4.5). The SSP2-45
scenario is chosen because mitigation technologies and policies are enacted
causing a general reduction in pollutants and \glspl{ghg}, including (eventual)
negative \cotwo emissions in some regions and sectors due to carbon capture and
sequestration and afforestation. A scenario in which negative emissions play a
role in mitigation strategies is particularly important because of the
sensitivity of key indicators, such as end-of-century radiative forcing (which
is used to estimate mean global temperature response), to the timing and
magnitude of net-zero and total negative \cotwo emissions. Therefore, these two
scenarios represent two contrasting cases in the use of a harmonization approach
and thus provide a case study as to its general applicability. Figure
\ref{fig:kyoto} shows the different trends of Kyoto Gases, a measure of
aggregate \gls{ghg} emissions, in each scenarios.

@@

We investigate the aggregate effect of harmonization with all necessary override
methods on total anthropogenic radiative forcing projections with the simple
carbon-cycle and climate model, MAGICC\footnote{MAGICC is a reduced complexity
  climate model which incorporates future trajectories of forcing agents (i.e.,
  emissions) to estimate future radiative forcing and mean global temperature
  response. We refer the reader to the
  \href{http://wiki.magicc.org/index.php?title=Model_Description}{MAGICC wiki}
  for a more in-depth description.} \cite{meinshausen2011emulating,
  meinshausen2011rcp}, for each harmonized and unharmonized scenario
respectively as shown in Figure \ref{fig:forcing}. We find that the change due
to harmonization is small, ranging between 1 and 2.5\% over the modeled time
period. Relative near-term differences persist in the mitigation case (SSP-4.5)
because differences in near-term emissions define to a larger degree the
longer-term forcing outcome due to the cumulative nature of long-lived climate
forcers like \cotwo.  The resulting difference in forcing in 2100 is 0.04
$\frac{\text{W}}{\text{m}^2}$ for SSP2-4.5 and 0.01
$\frac{\text{W}}{\text{m}^2}$ for SSP2-Ref, both of which are well within
acceptable tolerances (e.g., 0.75 $\frac{\text{W}}{\text{m}^2}$ defined for
\gls{smip} \cite{oneill_scenario_2016}). Thus harmonization is considered to
have a negligible effect on key long-term climate indicators.

>>>>

\texttt{aneris}, first made available in 2017, is available online at
\url{https://github.com/iiasa/aneris} as a free and open-source Python software
library (approximately 2000 lines of code). The \texttt{aneris} software was
developed by the lead author whose contact information is shown on the title
page of this manuscript. Documentation for the \texttt{aneris} Python package,
including software requirements, is available online at
\url{http://data.ene.iiasa.ac.at/gidden/aneris/}.

@@

\glspl{iam} are tools used to understand the complex interactions between
energy, economy, land use, water, and climate systems. \glspl{iam} provide
global projections of systemic change by dividing the world into a number of
representative regions (typically 10 to 30), the definition of which is distinct
for each model \cite{krey_global_2014}. Results from \glspl{iam} are integral in
a number of international studies, which notably include projections of climate
and energy futures. Recently, the \gls{iam} community has developed scenarios
based on the \glspl{ssp} \cite{Riahi2017153} which quantify a variety of
potential global futures. The \glspl{ssp} are designed to be used in research
that include \gls{esm} simulations, climate impact, adaptation and climate
mitigation studies \cite{oneill_new_2013, vuuren_new_2013}.

While \glspl{iam} are implemented in myriad ways\footnote{IAM models are
  numerous and have a long history in the scientific literature. Various IAMs
  have collaborated to produce community IAM documentation (available online:
  \url{http://themasites.pbl.nl/models/advance/index.php/ADVANCE_wiki}) which
  readers can access for a full treatment of model implementation and
  features.}, including simulation and optimization, the core inputs and outputs
are similar across different models. Modeling teams incorporate data on energy
systems, land use, economics, demographics and emissions sources and
concentrations, among other data, in order to provide a consistent starting
point for future projections. The models then provide estimates of future
trajectories of these variables under various socio-economic and technological
assumptions as well as proposed policy constraints, e.g., targets for future
\gls{ghg} emissions.

@@

When participating in intercomparison exercises in which a consistent historical
starting point is required (e.g., in \gls{cmip6}), model teams incorporate a
single, common historical data set through \textit{harmonization}. Harmonization
refers to the process of adjusting model results to match a selected historical
time series such that the resulting future trajectories are consistent with the
original modeled results and provide a smooth transition from the common
historical data. In the emissions context, this means that each individual
combination of model region, model sector, and emissions species must be
harmonized. Depending on the total number of model regions, sectors, and
emissions species, this can require the selection of thousands to
tens-of-thousands of harmonization methods.

@@

The remainder of this paper describes the methodology and implementation of the
harmonization software \code{aneris} \cite{matthew_gidden_2017_802832}, written
in the Python programming language (detailed documentation is available online
at \url{http://data.ene.iiasa.ac.at/gidden/aneris/}). Section \ref{sec:meths}
provides a detailed description of the underlying mathematical components of
\code{aneris} as well as the procedural workflow. A case study of applying the
automated harmonization mechanism on two example \gls{iam} scenarios, one with
emissions growth and another with emissions mitigation, is presented in Section
\ref{sec:results}. Finally, the general effectiveness and potential future
improvements on the automated methodology is discussed in Section
\ref{sec:future}.

@@

The goals of any scenario harmonization exercise are threefold: aligning model
results in the harmonization year to a common historical data source, faithfully
representing the original \gls{iam}’s internal consistency between the driver of
emissions (e.g. energy use) and emissions, and maintaining critical parameters
from the original scenario design. Any harmonization method achieves the first
goal by design. If the difference between the model base year and historical
values are small, considering the second and third goals leads to a method
choice that matches modeled drivers (e.g., a \code{ratio} method discussed in
Section \ref{sec:harm-methods}) and converges prior to the final model year. It
preserves the relationship between \gls{iam} output and emissions inventory in
the base year while also matching the original model output at some point in the
modeled time period. It furthermore maintains the consistency of the model's
usage of energy technology, volume of agricultural activities, and abatement
options with harmonized emissions trajectory.

@@

Accordingly, we have developed a \textit{decision tree} approach to
harmonization method choice, discussed in Section \ref{sec:tree}, in order to
balance each of these concerns and use cases in a robust, systematic,
reproducible, and transparent manner.

@@

\gls{iam} emission results are provided along temporal (normally half decade or
decade), spatial (i.e., model regions), emissions species, and sectoral
dimensions. Each individual temporal trajectory, i.e., unique combinations of
regions ($r$), species ($g$), and sectors ($s$), must be harmonized to the
initial modeling period. Given a model trajectory, $m_{r, g, s}(t)$, historical
values, $h_{r, g, s}(t)$, and model base year, $t_i$, a harmonized trajectory
needs to be calculated. The \textit{harmonization quality} of a trajectory,
i.e., how well a given harmonization algorithm performs, depends on a number of
factors. Of chief import is the faithful representation of the original
unharmonized trajectory as well as the representation of negative trajectories
(i.e., if a trajectory becomes negative, both the timing and total magnitude
should be as close as possible) which are of critical importance for cumulative
\cotwo calculations.

@@

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent \gls{ceds} and \gls{luc}
historical data has been performed. Figure \ref{fig:cov} shows the distribution
of \gls{luc} $c_v$s and non-\gls{luc} $c_v$s as determined for historical data
aggregated to the model regions of 5 different \glspl{iam} involved in the
\gls{ssp} process: AIM-CGE \cite{fujimori_ssp3:_2017},
IMAGE\cite{van_vuuren_energy_2017}, GCAM4\cite{calvin_ssp4:_2017},
MESSAGE-GLOBIOM\cite{fricko_marker_2017, message_globiom_2016}, and
REMIND-MAGPIE\cite{kriegler_fossil-fueled_2017}, each of which have varying
definitions of native model regions comprising different collections of
countries. Therefore, each data point comprising Figure \ref{fig:cov} represents
a realization of $c_v$ for a single combination of native model region, sector,
and emissions species\footnote{A full listing of all sectors and species is
  presented in the case-study discussion in Section \ref{sec:results}, Table
  \ref{tab:sp}}. A threshold value of $\tau_{c_v} = 20$ has been chosen based on
these observations as it optimally divides the two distributions. Importantly,
tails of the \gls{luc} and non-\gls{luc} overlap, thus there are both false
positives (~7\% of non-\gls{luc} trajectories) and false negatives (~10\% of
\gls{luc} trajectories). However, as any regional definition is model dependent
and thus any regional aggregation is possible an automated detection mechanism
is necessary.

@@

\begin{listing}[H]
\begin{minted}{python}
from aneris.tutorial import load_data
model, hist, driver = load_data()
for scenario in driver.scenarios():
    driver.harmonize(scenario)
harmonized, metadata = driver.harmonized_results()
\end{minted}
\caption{High-level user interaction with the \texttt{HarmonizationDriver} taken from the online \href{http://data.ene.iiasa.ac.at/gidden/aneris/tutorial.html}{tutorial}}
\label{lst:tutorial}
\end{listing}

@@

optional arguments:
  -h, --help            show this help message and exit
  --history HISTORY     Historical emissions in the base year.
  --regions REGIONS     Mapping of country iso-codes to native regions.
  --rc RC               Runcontrol YAML file (see
                        http://data.ene.iiasa.ac.at/gidden/aneris/config.html for
                        examples).
  --output_path OUTPUT_PATH
                        Path to use for output file names.
  --output_prefix OUTPUT_PREFIX
                        Prefix to use for output file names.
\end{minted}
\caption{The \gls{cli} help provided by the \code{aneris} package.}
\label{lst:cli}
\end{listing}

@@

Users are able to control the harmonization process via a number of options
(with examples provided
\href{http://data.ene.iiasa.ac.at/gidden/aneris/config.html}{online}). The primary mechanism
by which users control the process is by providing \textit{override} methods for
any combination of region and variable (i.e., sector and gas species).  In
practice, it may be possible that not all default methods chosen will provide
robust harmonized trajectories, especially if there is a significant difference
between historical and model values in the harmonization year, if there is
significant upward or downward movement in the model trajectory, or if there are
known discrepancies in sectoral definition between the \gls{iam} and historical data
source. In such cases, users can override default methods with their method of
choice and both the default method and override is reported in the resulting
metadata.

@@

In order to show a representative cross section of the performance of the
\code{aneris} harmonization procedure, we focus on the harmonization of results
of the \gls{iam} MESSAGE-GLOBIOM \cite{fricko_marker_2017,
  message_globiom_2016}. Harmonization results for two scenarios from the
\gls{ssp} scenario library\footnote{We refer the reader to the broad literature
  discussing the SSPs, e.g., \cite{Riahi2017153, oneill_new_2013,
    vuuren_new_2013,oneill_scenario_2016} for a more in depth discussion of the
  scenario architecture and design. Model results for various SSP scenarios are
  available online at \url{https://tntcat.iiasa.ac.at/SspDb}} are presented
here. We use the SSP2-reference\cite{fricko_marker_2017,Rao2017346}, or ``middle
of the road'', scenario (referred to as SSP2-Ref) as an example because
MESSAGE-GLOBIOM is the marker scenario\footnote{The ``marker'' scenario concept
  is used to designate the archetype scenario used as a reference within each
  SSP scenario family. See \cite{Riahi2017153} and \cite{vuuren_new_2013} for a
  more lengthy description.} for this \gls{ssp}. This SSP2 scenario lies between
two \glspl{rcp}, 6 and 8.5, with a radiative forcing\footnote{Radiative forcing
  in this context is the energy imbalance at the top of the atmosphere caused by
  anthropogenic influences relative to a pre-industrial reference point. Higher
  radiative forcing leads to larger global changes, such as surface
  temperature. We refer the reader to \cite{IPCCWG1PhysicalStocker2013} for a
  more detailed discussion.} level of approximately 6.5 Wm$^{-2}$. We
additionally present the results for the SSP2-based mitigation scenario leading
to a radiative forcing of 4.5 Wm$^{-2}$ (referred to as SSP2-4.5). The SSP2-45
scenario is chosen because mitigation technologies and policies are enacted
causing a general reduction in pollutants and \glspl{ghg}, including (eventual)
negative \cotwo emissions in some regions and sectors due to carbon capture and
sequestration and afforestation. A scenario in which negative emissions play a
role in mitigation strategies is particularly important because of the
sensitivity of key indicators, such as end-of-century radiative forcing (which
is used to estimate mean global temperature response), to the timing and
magnitude of net-zero and total negative \cotwo emissions. Therefore, these two
scenarios represent two contrasting cases in the use of a harmonization approach
and thus provide a case study as to its general applicability. Figure
\ref{fig:kyoto} shows the different trends of Kyoto Gases, a measure of
aggregate \gls{ghg} emissions, in each scenarios.

@@

We investigate the aggregate effect of harmonization with all necessary override
methods on total anthropogenic radiative forcing projections with the simple
carbon-cycle and climate model, MAGICC\footnote{MAGICC is a reduced complexity
  climate model which incorporates future trajectories of forcing agents (i.e.,
  emissions) to estimate future radiative forcing and mean global temperature
  response. We refer the reader to the MAGICC wiki (available online:
  \url{http://wiki.magicc.org}) for a more in-depth description.}
\cite{meinshausen2011emulating, meinshausen2011rcp}, for each harmonized and
unharmonized scenario respectively as shown in Figure \ref{fig:forcing}. We find
that the change due to harmonization is small, ranging between 1 and 2.5\% over
the modeled time period. Relative near-term differences persist in the
mitigation case (SSP-4.5) because differences in near-term emissions define to a
larger degree the longer-term forcing outcome due to the cumulative nature of
long-lived climate forcers like \cotwo.  The resulting difference in forcing in
2100 is 0.04 $\frac{\text{W}}{\text{m}^2}$ for SSP2-4.5 and 0.01
$\frac{\text{W}}{\text{m}^2}$ for SSP2-Ref, both of which are well within
acceptable tolerances (e.g., 0.75 $\frac{\text{W}}{\text{m}^2}$ defined for
\gls{smip} \cite{oneill_scenario_2016}). Thus harmonization is considered to
have a negligible effect on key long-term climate indicators.
